---
title: "Using PLO(SC)²"
output: html_notebook
---

This R-Markdown will guide you through an example analysis using PLO(SC)².

```{r}

library("Seurat")
library("PLOSC")

```


Reading in all files and automatically assigning sample names

```{r}

files <- Sys.glob("T:/scdata/covid_sc_cellranger/h5files/*.h5")
inputMatrices = readH5Files(files, sample_element=5, sample_processor=function(x){return(substr(x, 1, 10))})

```

Transforms the inputMatrices into Seurat objects, and preprocesses these for 3000 highly variable genes.

```{r}

objlist.raw = toObjList(inputMatrices, patternList_human(), 3000)

```

The input matrices (in the example case these contain all cells) are filtered: 100 < nFeature_RNA < 6000, nCount_RNA > 500 and percent_mt < 7.

```{r}

objlist = scatterAndFilter(objlist.raw, nfeature_rna.lower=100, nfeature_rna.upper=6000, ncount_rna.lower=500, percent_mt.upper=7, plot=FALSE, plot.after=TRUE)

```

Number of remaining cells per experiment

```{r}

print("cells per experiment")
print(mapply(sum, lapply(objlist, function(x) {dim(x)[2]})))
print("total cells")
print(sum(mapply(sum, lapply(objlist, function(x) {dim(x)[2]}))))

```

The raw input files are not needed anymore and deleted to save some RAM.

```{r}

objlist.raw = NULL
inputMatrices = NULL

```

The different libraries are meant to be integrated/batch removed. This step is prepared here, by first estimating cell-cycle per cell, calculating 3000 highly variable features and regressing out the cell-cycle genes in 3000 scaled genes (which are also those used for the integrating the datasets.)

```{r}

finalList = prepareIntegration(objlist, cc.use.genes = cc.genes, nfeatures.variable = 3000, nfeatures.scale=3000, run.parallel=FALSE)

```

This step performs the actual integration using the features (genes) determined in the previous command, and using the rpca method. On some R installations running RunPCA in parallel stalls R, hence it can be switched off here. No further assay (e.g. CITE-seq) is to be integrated, hence add.do is set to False.

```{r}

integratedList_sample = performIntegration(finalList$data, "libintegration", features.integration = finalList$features, gex.method.normalization="LogNormalize", gex.method.integration="rpca", add.do=FALSE, run.parallel=TRUE)

```

The ElbowPlot helps us to estimate how many PCs are needed for dimensional reduction (e.g. UMAP).

```{r}

ElbowPlot(integratedList_sample$integrated, ndims=30, reduction = "igpca")

```

The preprocessIntegrated function calculates a UMAP representation of the data and clusters the cells.

```{r}

obj.integrated = preprocessIntegrated(integratedList_sample$integrated, "integratedgex", "libintegration",  resolution=0.5, num.pcs=10, dim.reduction="igpca", with.hto=FALSE)


```

This removes the UMAP calculated during the performIntegration-step.

```{r}
obj.integrated@reductions$ig.umap = NULL
```

Showing the calculated UMAP representation.

```{r}
DimPlot(obj.integrated, label=T)
```

```{r}

df = read.table("samples2condition.tsv", header = TRUE)
df[order(df$sampleid),]

```

We now add meta-data to our object for further processing.

```{r}

annotateList.sample_condition = list(
  list(name="SYMPTOMATIC", selector="^20094_0001"),
  list(name="SYMPTOMATIC", selector="^20094_0002"),
  list(name="SYMPTOMATIC", selector="^20094_0003"),
  list(name="ASYMPTOMATIC", selector="^20094_0004"),
  list(name="ASYMPTOMATIC", selector="^20094_0005"),
  list(name="ASYMPTOMATIC", selector="^20094_0006"),
  list(name="SYMPTOMATIC", selector="^20094_0007"),
  list(name="SYMPTOMATIC", selector="^20094_0008"),
  list(name="SYMPTOMATIC", selector="^20094_0009"),
  list(name="CONTROL", selector="^20094_0012")

)

obj.integrated = annotateByCellnamePattern( obj.integrated, "condition", annotateList.sample_condition, order=c("CONTROL", "ASYMPTOMATIC", "SYMPTOMATIC"))

annotateList.sample_condition = list(
  list(name="TP 1", selector="^20094_0001"),
  list(name="TP 2", selector="^20094_0002"),
  list(name="TP 3", selector="^20094_0003"),
  list(name="TP 1", selector="^20094_0004"),
  list(name="TP 2", selector="^20094_0005"),
  list(name="TP 3", selector="^20094_0006"),
  list(name="TP 1", selector="^20094_0007"),
  list(name="TP 2", selector="^20094_0008"),
  list(name="TP 3", selector="^20094_0009"),
  list(name="Ctrl", selector="^20094_0012")
)

obj.integrated = annotateByCellnamePattern( obj.integrated, "tp", annotateList.sample_condition, order=c("Ctrl", "TP 1", "TP 2", "TP 3"))


```

The makeUMAPPlot functin draws a 2D UMAP array, which is a good tool to visualize a time-series experiment across several conditions.

```{r, fig.width=24, fig.height=20}
p=makeUMAPPlot(obj.integrated, dim1 = "condition", dim2="tp", group.by="idents", downsample = FALSE)
save_plot(p, "split_dimplot_downsample_no", 24, 20)

```

The downsample-options reduces the number of shown cells in all plots to the same number (the minimum of all).

```{r, fig.width=24, fig.height=20}
p=makeUMAPPlot(obj.integrated, dim1 = "condition", dim2="tp", group.by="idents", downsample = TRUE)
save_plot(p, "split_dimplot_downsample", 24, 20)


```


Next we perform marker-gene analysis (DE genes per cluster). The makeDEResults function calls FindAllMarkers, but enhances these results with a summary statistic per gene.

```{r}

dir.create("libintegration/de")

exprdfTT = makeDEResults(obj.integrated, assay="RNA", test="t")
write.table(exprdfTT, "libintegration/de/expr_test_t.tsv", sep="\t", row.names=F, quote = F)
writexl::write_xlsx(exprdfTT, "libintegration/de/expr_test_t.xlsx")
print(head(exprdfTT))

```


Next we calculate a module score for a set of interferon-stimulating genes, which are analysed in the following plots.

```{r}
DefaultAssay(obj.integrated)= "RNA"

isgGenes = unique(c("MT2A", "ISG15", "LY6E", "IFIT1", "IFIT2", "IFIT3", "IFITM1", "IFITM3", "IFI44L", "IFI6", "MX1", "IFI27",  "IFI44L", "RSAD2", "SIGLEC1", "IFIT1"))

obj.integrated = AddModuleScore(obj.integrated, features =  isgGenes, name = "isg_score_small")

```

Let's briefly visualize the gene signature:

```{r}
FeaturePlot(obj.integrated, "isg_score_small1")

```

```{r}
obj.integrated$cond_tp = paste(obj.integrated$condition, obj.integrated$tp, sep=" ")
```

The splitFeaturePlot is a separate FeaturePlot for each panel defined by the split.by clause. All panels have the same limits and are thus easily comparable.

```{r, fig.width=30, fig.height=15}
p=splitFeaturePlot(obj.integrated, "isg_score_small1", split.by="cond_tp", title="FeaturePlot", ncol=4)
save_plot(p, "split_featureplot_isg", 30, 15)

```

The comparativeVioBoxPlot combines a BoxPlot with a Violin Plot and performs statistical tests within each group (between the subsets defined by split.by).

```{r fig.width=12, fig.height=8}

p=comparativeVioBoxPlot(obj.integrated, "isg_score_small1", group.by="condition", split.by="tp", yStepIncrease = 1, verbose=FALSE)
save_plot(p, "comparative_violin_isg", 12, 8)

```

The plot is also customizable in terms of colors.

```{r fig.width=12, fig.height=8}

dsrCols = list("Ctrl"="steelblue4", "TP 1"="darkred", "TP 2"="chocolate2", "TP 3"="darkseagreen3")

p=comparativeVioBoxPlot(obj.integrated, "isg_score_small1", group.by="condition", split.by="tp", yStepIncrease = 1, verbose=FALSE, dsrCols = dsrCols, boxplot_grey = "ivory4")
save_plot(p, "comparative_violin_isg_custom", 12, 8)

```

```{r}
unique(obj.integrated$condition)
```
And also works when only one major group is available.

```{r}

dsrCols = list("Ctrl"="steelblue4", "TP 1"="darkred", "TP 2"="chocolate2", "TP 3"="darkseagreen3")

comparativeVioBoxPlot(subset(obj.integrated, condition=="ASYMPTOMATIC"), "isg_score_small1", group.by="condition", split.by="tp", yStepIncrease = 1, verbose=FALSE, dsrCols = dsrCols, boxplot_grey = "ivory4")

```

```{r}
obj.integrated = Seurat::ScaleData(obj.integrated)
obj.integrated

```

```{r}
unique(obj.integrated$idents)
```
Single heatmap plot of the isgGenes per cluster. Scaled globally from ScaleData.

```{r}

plotGOIs = list("1"=list("clusters"=c("0"), "genes"=isgGenes))

p=enhancedHeatMap(obj.integrated, plotGOIs, group.by="idents", include_all_clusters = TRUE, title="Heatmap of ISG Genes", scale.by="GLOBAL")
save_plot(p, "heatmap_isg_genes", 12, 8, save.data = FALSE)
```
Single heatmap plot of the isgGenes per cluster. Scaled by all shown values in the heatmap.

```{r}

plotGOIs = list("1"=list("clusters"=c("0"), "genes"=isgGenes))

p=enhancedHeatMap(obj.integrated, plotGOIs, group.by="idents", include_all_clusters = TRUE, title="Heatmap of ISG Genes", scale.by="ALL")
save_plot(p, "heatmap_isg_genes_ALL", 12, 8, save.data = FALSE)


```
Split heatmap plot, grouping the cells per cluster and split the results per condition. Scaled globally from ScaleData

```{r fig.width=30, fig.height=10}

p=makeComplexExprHeatmapSplit(obj.integrated, plotGOIs, group.by="idents", split.by="condition", include_all_clusters = TRUE, title="Heatmap of ISG Genes", scale.by="GLOBAL")
save_plot(p, "heatmap_split_isg_genes", 30, 10, save.data = FALSE)


```

Split heatmap plot, grouping the cells per cluster and split the results per condition. Scaled by all shown values (over all 3 groups).

```{r fig.width=30, fig.height=10}

p=makeComplexExprHeatmapSplit(obj.integrated, plotGOIs, group.by="idents", split.by="condition", include_all_clusters = TRUE, title="Heatmap of ISG Genes", scale.by="ALL")
save_plot(p, "heatmap_split_isg_genes_ALL", 30, 10, save.data = FALSE)


```

Split heatmap plot, grouping the cells per cluster and split the results per condition. Scaled per group/heatmap.

```{r fig.width=30, fig.height=10}


p=makeComplexExprHeatmapSplit(obj.integrated, plotGOIs, group.by="idents", split.by="condition", include_all_clusters = TRUE, title="Heatmap of ISG Genes", scale.by="GROUP")
save_plot(p, "heatmap_split_isg_genes_GROUP", 30, 10, save.data = FALSE)


```

This prepares the layout for the enhancedDotPlot. Per panel one entry in the plot elems with selected cells and a label is required.

```{r, fig.height=12, fig.width=16}

cells.control = cellIDForClusters(obj.integrated, "condition", c("CONTROL"))
cells.symptomatic = cellIDForClusters(obj.integrated, "condition", c("SYMPTOMATIC"))
cells.asymptomatic = cellIDForClusters(obj.integrated, "condition", c("ASYMPTOMATIC"))

plotElems = list()
plotElems[["Control"]] = list(cells=intersect(cells.control, cells.control), label="Control")
plotElems[["Symptomatic"]] = list(cells=intersect(cells.symptomatic,cells.symptomatic), label="Symptomatic")
plotElems[["Asymptomatic"]] = list(cells=intersect(cells.asymptomatic,cells.asymptomatic), label="Asymptomatic")

```


enhancedDotPlot of isg genes, scaled per all shown values.


```{r, fig.height=12, fig.width=16}

p=enhancedDotPlot(obj.integrated, plotElems, featureGenes = isgGenes, group.by="idents", title="DotPlot of ISG Genes", scale.by="ALL", rotate.x = T)
save_plot(p, "edotplot_split_isg_genes_ALL", 30, 10, save.data = FALSE)

```

enhancedDotPlot of isg genes, scaled globally (from ScaleData).

```{r, fig.height=12, fig.width=16}

p=enhancedDotPlot(obj.integrated, plotElems, featureGenes = isgGenes, group.by="idents", title="DotPlot of ISG Genes", scale.by="GLOBAL", rotate.x = T)
save_plot(p, "edotplot_split_isg_genes_GLOBAL", 30, 10, save.data = FALSE)


```

enhancedDotPlot of isg genes, scaled per shown group.

```{r, fig.height=12, fig.width=16}

p=enhancedDotPlot(obj.integrated, plotElems, featureGenes = isgGenes, group.by="idents", title="DotPlot of ISG Genes", scale.by="GROUP",rotate.x = T)
save_plot(p, "edotplot_split_isg_genes_GROUP", 30, 10, save.data = FALSE)


```

enhancedDotPlot of isg genes, scaled per feature (across all 3 shown groups).

```{r, fig.height=12, fig.width=16}

p=enhancedDotPlot(obj.integrated, plotElems, featureGenes = isgGenes, group.by="idents", title="DotPlot of ISG Genes", scale.by="FEATURE", rotate.x=TRUE)
save_plot(p, "edotplot_split_isg_genes_FEATURE", 30, 10, save.data = FALSE)


```

Compare symptomatic vs asymptomatic in all clusters (at TP1).

```{r}

cells.tp1 = cellIDForClusters(obj.integrated, "tp", c("TP 1"))

comparison.sympt_asympt = compareCellsByCluster(obj.integrated, intersect(cells.symptomatic, cells.tp1), intersect(cells.asymptomatic, cells.tp1), "symptomatic", "asymptomatic", outfolder="./de", group.by="idents", heatmap.plot=TRUE)

```

Visualize the above comparison using volcano plots.

```{r}

makeVolcanos(comparison.sympt_asympt, "Comparison Symptomatic vs. Asymptoamtic", "./de_volcanos/sympt_asympt_tp1", FCcutoff=0.25)

```

Perform enrichment analysis on the above DE comparison.

```{r}

gseResults = performEnrichtmentAnalysis(obj.integrated, comparison.sympt_asympt, "human", "./gse_results.rds")


```

Let's visualize the enrichment results using clusterProfiler.

```{r}

makeEnrichmentPlots( gseResults, outfolder = "enrichment_plots/" )


```

```{r}
print(sessionInfo())
```

```{r}
save.image("vignette_using_plosc.RData")
```

